Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-19 paper/sections/methodology.rmd
incorporated Cynthia's comments

In this section we first set out a generalized framework for gradient-based counterfactual search in Section \@ref(method-general) that encapsulates the various individual recourse methods we have chosen to use in our experiments. We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective.

@@

We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method - Wachter; secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work. In particular, we include include the following generators in our simulation experiments below: **REVISE** [@joshi2019towards], **CLUE** [@antoran2020getting], **DiCE** [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating]. Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models. We hypothesize that generating more **meaningful** counterfactuals should mitigate the endogenous dynamics illustrated in Figure \@ref(fig:poc) in Section \@ref(intro). This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data. All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.

@@

Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is the stacked $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space. In Wachter, the state space is the feature space: $f$ is just the identity function and the number of counterfactuals $K$ is equal to one. Both REVISE and CLUE search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$. In this case $f$ in Equation \@ref(eq:general) corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space. Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019towards].

CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, $M$. To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models. The approach proposed by @schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates. The authors show that in this setting the cost function $\text{cost}(\cdot)$ in Equation \@ref(eq:general) is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Schut et al. [@schut2021generating] also show that by maximizing the predicted probability of $x^\prime$ being assigned to target class $y^*$ we also implicitly minimize predictive entropy - as in CLUE. In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as **Latent Space** generators^[In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category. They largely differ with respect to the chosen generative model: for example, the generator proposed by @dombrowski2021diffeomorphic relies on normalizing flows.].

Finally, DiCE [@mothilal2020explaining] distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals. Wachter et al. [@wachter2017counterfactual] show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual)^[Note, in fact, that \@ref(eq:general) naturally lends itself to that idea: setting $K$ to some value greater than one and using the Wachter objective essentially boils down to computing multiple counterfactuals in parallel. Here, $yloss(\cdot)$ is first broadcasted over elements of $\mathbf{s}^\prime$ and then aggregated. This is exactly how counterfactual search is implemented in `CounterfactualExplanations.jl`.]. In @mothilal2020explaining diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function $\text{cost}(\mathbf{s}^\prime)$ and thereby produce counterfactuals $s_1, ... , s_K$ that look as different from each other as possible. The implementation of DiCE in the our library of choice - `CounterfactualExplanations.jl` - uses that exact approach. It is worth noting that for $k=1$, DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \@ref(eq:general).

## ... towards collective recourse

>>>>

In this section we first set out a generalized framework for gradient-based counterfactual search that encapsulates the various individual recourse methods we have chosen to use in our experiments (Section \@ref(method-general)). We then introduce the notion of a hidden external cost in algorithmic recourse and extend the existing framework to explicitly address this cost in the counterfactual search objective (Section \@ref(method-collective)).

@@

We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method (Wachter); secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work. In particular, we include the following generators in our simulation experiments below: **REVISE** [@joshi2019realistic], **CLUE** [@antoran2020getting], **DiCE** [@mothilal2020explaining] and a greedy approach that relies on probabilistic models [@schut2021generating]. Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models. We hypothesize that generating more **meaningful** counterfactuals should mitigate the endogenous dynamics illustrated in Figure \@ref(fig:poc) in Section \@ref(intro). This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data. All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.

@@

Here $\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K$ is the stacked $K$-dimensional array of counterfactual states  and $f: \mathcal{S} \mapsto \mathcal{X}$ maps from the counterfactual state space to the feature space. In Wachter, the state space is the feature space: $f$ is just the identity function and the number of counterfactuals $K$ is equal to one. Both REVISE and CLUE search counterfactuals in some latent embedding $S \subset \mathcal{S}$ instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of $X$. In this case, $f$ in Equation \@ref(eq:general) corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space. Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP [@joshi2019realistic].

CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, $M$. To quantify predictive uncertainty, Antoran et al. [@antoran2020getting] rely on entropy estimates for probabilistic models. The greedy approach proposed by Schut et al. @schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates. The authors show that in this setting the cost function $\text{cost}(\cdot)$ in Equation \@ref(eq:general) is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Schut et al. [@schut2021generating] also show that by maximizing the predicted probability of $x^\prime$ being assigned to target class $y^*$, we also implicitly minimize predictive entropy (as in CLUE). In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as **Latent Space** generators^[In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category. They largely differ with respect to the chosen generative model: for example, the generator proposed by @dombrowski2021diffeomorphic relies on normalizing flows.].

Finally, DiCE [@mothilal2020explaining] distinguishes itself from all other generators considered here in that it aims to generate a diverse set of $K>1$ counterfactuals. Wachter et al. [@wachter2017counterfactual] show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual)^[Note, in fact, that \@ref(eq:general) naturally lends itself to that idea: setting $K$ to some value greater than one and using the Wachter objective essentially boils down to computing multiple counterfactuals in parallel. Here, $yloss(\cdot)$ is first broadcasted over elements of $\mathbf{s}^\prime$ and then aggregated. This is exactly how counterfactual search is implemented in `CounterfactualExplanations.jl`.]. In @mothilal2020explaining diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function $\text{cost}(\mathbf{s}^\prime)$ and thereby produce counterfactuals $s_1, ... , s_K$ that look as different from each other as possible. The implementation of DiCE in our library of choice---`CounterfactualExplanations.jl`---uses that exact approach. It is worth noting that for $k=1$, DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \@ref(eq:general).

## ... towards collective recourse {#method-collective}

### 2022-08-19 paper/sections/methodology_2.rmd
incorporated Cynthia's comments

In the following we describe the framework we propose for modeling and analysing endogenous macrodynamics in Algorithmic Recourse. We introduce this framework with the ambition to shed light on the following research questions:

@@

Below we first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md) - the Julia package we introduced earlier. The remainder of this section then introduces various evaluation metrics that can be used to benchmark different counterfactual generators with respect to how they perform in the dynamic setting.

@@

The dynamics illustrated in Figure \@ref(fig:poc) in Section \@ref(intro) were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice. We begin in the static setting at time time $t=0$: firstly, we have some binary classifier $M$ that was pre-trained on data $\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1$, where $\mathcal{D}_0$ and $\mathcal{D}_1$ denote samples in the non-target and target class, respectively; secondly, we generate recourse for a random batch of $B$ individuals in the non-target class ($\mathcal{D}_0$). Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse^[To keep notation simple we have also restricted ourselves to binary classification here, but `AlgorithmicRecourseDynamics.jl` can also be used for multi-class problems.].

In order to simulate the dynamic process, we suppose that the model $M$ is retrained following the actual implementation of recourse in time $t=0$. Following the update to the model, we assume that at time $t=1$ recourse is generated for yet another random subset of individuals in the non-target class. This process is repeated for a number of time periods $T$. To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population. The experimental setup is summarized in Algorithm \ref{algo-experiment}

@@

Algorithm \ref{algo-experiment} summarizes the proposed simulation experiment for a given dataset $\mathcal{D}$, model $M$ and generator $G$, but naturally we are interested in comparing simulation outcomes for different sources of data, models and generators. The framework we have built facilitates this, making use of multi-threading in order to speed up computations. Holding the initial model and dataset constant the experiments are run for all generators, since our primary concern is to benchmark different recourse methods. To ensure that each generator is faced with the exact same initial conditions in each round $t$, the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all individuals in the non-target class across all experiments $\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J$ where $J$ is the total number of generators.

@@

where $X=\{x_1,...,x_m\}$, $\tilde{X}=\{\tilde{x}_1,...,\tilde{x}_n\}$ represent independent and identically distributed samples drawn from probability distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$ respectively @gretton2012kernel. MMD is a measure of the distance between the kernel mean embeddings of $\mathcal{X}$ and $\mathcal{\tilde{X}}$ in a Reproducing Kernel Hilbert Space, $\mathcal{H}$ [@berlinet2011reproducing]. An important consideration is the choice of the kernel function $k(\cdot,\cdot)$. In our implementation we make use of Gaussian kernel with a constant length-scale parameter of $0.5$. As the Gaussian kernel captures all moments of distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$, we have that $MMD(X,\tilde{X})=0$ if and only if $X=\tilde{X}$.

The evaluation metric in Equation \@ref(eq:mmd) is computed after every round $t=1,...,T$ of the experiment. To assess the statistical significance of the observed shifts under the null hypothesis that samples $X$ and $\tilde{X}$ were drawn from the same probability distribution, we follow @arcones1992bootstrap. To that end, we combine the two samples and generate a large number of permutations of $X + \tilde{X}$. Then, we split the permuted data into two new samples $X^\prime$ and $\tilde{X}^\prime$ having the same size as the original samples. Then under the null hypothesis we should have that $MMD(X^\prime,\tilde{X}^\prime)$ be approximately equal to $MMD(X,\tilde{X})$. The corresponding $p$-value can then be calculated by counting how these two quantities are not equal.

@@

As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time $t$ following @upadhyay2021towards. We define $\Delta=||\theta_{t+1}-\theta_{t}||^2$, that is the euclidean distance between the vectors of parameters before and after retraining the model $M$. We shall refer to this baseline metric simply as **Perturbations**.

@@

As an alternative to PP MMD we use a pseudo-distance for the **Disagreement Coefficient** (Disagreement). This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M^\prime(x))$, that is the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample. Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries. In our context, this metric quantifies the overlap between the initial model (trained before the application of recourse) and the updated model. A Disagreement Coefficient unequal to zero is indicative of a model shift. The opposite is not true: even if the Disagreement Coefficient is equal to zero a model shift may still have occured. This is one reason for why PP MMD is our our preferred metric.

>>>>

In the following we describe the framework we propose for modeling and analyzing endogenous macrodynamics in Algorithmic Recourse. We introduce this framework with the ambition to shed light on the following research questions:

@@

Below we first describe the basic simulations that were generated to produce the findings in this work and also constitute the core of [`AlgorithmicRecourseDynamics.jl`](https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md)---the Julia package we introduced earlier. The remainder of this section then introduces various evaluation metrics that can be used to benchmark different counterfactual generators with respect to how they perform in the dynamic setting.

@@

The dynamics illustrated in Figure \@ref(fig:poc) in Section \@ref(intro) were generated through a simple experiment that aims to simulate the process of Algorithmic Recourse in practice. We begin in the static setting at time time $t=0$: firstly, we have some binary classifier $M$ that was pre-trained on data $\mathcal{D}=\mathcal{D}_0 \cup \mathcal{D}_1$, where $\mathcal{D}_0$ and $\mathcal{D}_1$ denote samples in the non-target and target class, respectively; secondly, we generate recourse for a random batch of $B$ individuals in the non-target class ($\mathcal{D}_0$). Note that we focus our attention on classification problems, since classification poses the most common practical use-case for Algorithmic Recourse^[To keep notation simple, we have also restricted ourselves to binary classification here, but `AlgorithmicRecourseDynamics.jl` can also be used for multi-class problems.].

In order to simulate the dynamic process, we suppose that the model $M$ is retrained following the actual implementation of recourse in time $t=0$. Following the update to the model, we assume that at time $t=1$ recourse is generated for yet another random subset of individuals in the non-target class. This process is repeated for a number of time periods $T$. To get a clean read on endogenous dynamics we keep the total population of samples closed: we allow existing samples to move from factual to counterfactual states, but do not allow any entirely new samples to enter the population. The experimental setup is summarized in Algorithm \ref{algo-experiment}.

@@

Algorithm \ref{algo-experiment} summarizes the proposed simulation experiment for a given dataset $\mathcal{D}$, model $M$ and generator $G$, but naturally we are interested in comparing simulation outcomes for different sources of data, models and generators. The framework we have built facilitates this, making use of multi-threading in order to speed up computations. Holding the initial model and dataset constant, the experiments are run for all generators, since our primary concern is to benchmark different recourse methods. To ensure that each generator is faced with the exact same initial conditions in each round $t$, the candidate batch of individuals from the non-target class is randomly drawn from the intersection of all individuals in the non-target class across all experiments $\left\{\textsc{Experiment}(M,\mathcal{D},G)\right\}_{j=1}^J$ where $J$ is the total number of generators.

@@

where $X=\{x_1,...,x_m\}$, $\tilde{X}=\{\tilde{x}_1,...,\tilde{x}_n\}$ represent independent and identically distributed samples drawn from probability distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$ respectively @gretton2012kernel. MMD is a measure of the distance between the kernel mean embeddings of $\mathcal{X}$ and $\mathcal{\tilde{X}}$ in a Reproducing Kernel Hilbert Space, $\mathcal{H}$ [@berlinet2011reproducing]. An important consideration is the choice of the kernel function $k(\cdot,\cdot)$. In our implementation we make use of a Gaussian kernel with a constant length-scale parameter of $0.5$. As the Gaussian kernel captures all moments of distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$, we have that $MMD(X,\tilde{X})=0$ if and only if $X=\tilde{X}$.

The evaluation metric in Equation \@ref(eq:mmd) is computed after every round $t=1,...,T$ of the experiment. To assess the statistical significance of the observed shifts under the null hypothesis that samples $X$ and $\tilde{X}$ were drawn from the same probability distribution, we follow @arcones1992bootstrap. To that end, we combine the two samples and generate a large number of permutations of $X + \tilde{X}$. Then, we split the permuted data into two new samples $X^\prime$ and $\tilde{X}^\prime$ having the same size as the original samples. Then under the null hypothesis, we should have that $MMD(X^\prime,\tilde{X}^\prime)$ be approximately equal to $MMD(X,\tilde{X})$. The corresponding $p$-value can then be calculated by counting how these two quantities are not equal.

@@

As our baseline for quantifying model shifts we measure perturbations to the model parameters at each point in time $t$ following @upadhyay2021robust. We define $\Delta=||\theta_{t+1}-\theta_{t}||^2$, that is the euclidean distance between the vectors of parameters before and after retraining the model $M$. We shall refer to this baseline metric simply as **Perturbations**.

@@

As an alternative to PP MMD, we use a pseudo-distance for the **Disagreement Coefficient** (Disagreement). This metric was introduced in @hanneke2007bound and estimates $p(M(x) \neq M^\prime(x))$, that is, the probability that two classifiers do not agree on the predicted outcome for a randomly chosen sample. Thus, it is not relevant whether the classification is correct according to the ground truth, but only whether the sample lies on the same side of the two respective decision boundaries. In our context, this metric quantifies the overlap between the initial model (trained before the application of recourse) and the updated model. A Disagreement Coefficient unequal to zero is indicative of a model shift. The opposite is not true: even if the Disagreement Coefficient is equal to zero a model shift may still have occured. This is one reason for why PP MMD is our our preferred metric.

### 2022-08-19 paper/sections/mitigation.rmd
incorporated Cynthia's comments

We proposed and test a number of simple mitigation strategies. All of them essentially boil down to one simple principle: to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process. This principle is really at the core of Latent Space generators, and hence it is not surprising that we have found these types of generators to perform comparably well in the previous section. But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational burden and - perhaps more importantly - their performance hinges on the performance of said generative models. Fortunately, it turns out that we can use a number of other, much simpler strategies, which we will discuss now.

@@

The most obvious and trivial mitigation strategy is to simply choose a higher decision threshold $\gamma$. This threshold determines when a proposed counterfactual should be considered as valid. Under $\gamma=0.5$, counterfactuals will end up near the decision boundary by construction. Since this is the region of maximal aleotoric uncertainty, the classifier is bound to be thrown off. By simply setting a more conservative decision threshold, we can avoid this issue to some extent. A drawback of this approach is that a classifier with high decisiveness may classify samples with high confidence even far away from the training data.

## Classifier Preserving ROAR

Another strategy draws inspiration from ROAR @upadhyay2021towards: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual $x^\prime$ at given parameter values. Recall that $\text{extcost}(\cdot)$ denotes what we had defined as the external cost in Equation \@ref(eq:collective). Then formally we let

@@

for each counterfactual $k$ where $l$ denotes the loss function used to train $M$. This approach, which we shall refer to as **ClapROAR**, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss. It is closely linked to the idea of choosing higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers. It also makes the private vs. external cost trade-off more explicit and hence manageable.

@@

Yet another strategy simply extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average: $\bar{x}$:

@@

Once again we can putting this in the context of Equation \@ref(eq:collective), the former penalty can be thought of here as the private cost incurred by the individual, while the latter reflects the external cost incurred by other individuals. Higher choices of $\lambda_2$ relative to $\lambda_1$ will lead counterfactuals to gravitate towards the specified point $\bar{x}$ in the target domain. In the remainder of this paper we will therefore refer to this approach as **Gravitational** generator, when we investigate its usefulness for mitigating endongenous macrodynamics^[Note that despite the naming convention our goal here is not to provide yet another counterfactual generator, but merely investigate the most simple penalty we can think of with respect to its effectiveness.].

Figure \@ref(fig:mitigation) shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClapROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Gravitational generator and ClapROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].

@@

# Generators:
generators = Dict(
    "Generic (γ=0.5)" => GenericGenerator(decision_threshold=0.5),
    "Generic (γ=0.9)" => GenericGenerator(decision_threshold=0.9),
    "Gravitational" => GravitationalGenerator(),
    "ClapROAR" => ClapROARGenerator()
)

@@

Our findings indicate that all three mitigation strategies are at least at par with Latent Space generators with respect to their effectiveness at mitigating domain and model shifts. Figure \@ref(fig:mitigate-results) presents a subset of the evaluation metrics for our synthetic data with overlapping classes. The top row in Figure \@ref(fig:mitigate-results) indicates that while domain shifts are of roughly the same magnitude for both Wachter and Latent Space generators, our proposed strategies effectively mitigate these shifts. ClapROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts. As evident from the middle row in Figure \@ref(fig:mitigate-results) model shifts can also be reduced: for the deep ensemble Latent Space search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective. The same overall pattern can be observed for out-of-sample model performance. With respect to the other synthetic datasets we found the following: for the Moons dataset the emerging patterns are largely the same, but the estimated model shifts are insignificant as noted earlier; the same holds for the Circles dataset, but there is no significant reduction in model performance for our neural networks; in the case of linearly separable data we find the Gravitational generator to be most effective at mitigating shifts.

@@

An interesting finding is also that the proposed strategies can have a complementary effect when used in combination with Latent Space generators. In experiments we conducted on the synthetic data, the apparent benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClapROAR. In Figure \@ref(fig:mitigate-latent-results) the conventional Latent Space generator with $\gamma=0.5$ serves as our baseline. Evidently, being more conservative or using one of our proposed penalties decreases the estimated domain and model shifts.

@@

Finally, Figure \@ref(fig:mitigate-real-world-results) shows the results for our real-world data. We note that for both California Housing and Credit Default data our proposed strategies do have an attenuating effect on both models shifts and performance deterioration, even though the estimated effects are somewhat less striking than for the synthetic data in \@ref(fig:mitigate-results)^[Estimated domain shifts (not shown) were largely insubstantial, as in Figure \@ref(fig:real) in the previous section.]. Still, both ClapROAR and Gravitational reduce the negative impact on out-of-sample model performance by around 25 percent from roughly 20 percentage points for the baseline approach to just 15 percentage points. For the GMSC dataset we observe no notable differences.

>>>>

We proposed and test a number of simple mitigation strategies. All of them essentially boil down to one simple principle: to avoid substantial domain and model shifts, the generated counterfactuals should comply as much as possible with the true data generating process. This principle is really at the core of Latent Space generators, and hence it is not surprising that we have found these types of generators to perform comparably well in the previous section. But as we have mentioned earlier, generators that rely on separate generative models carry an additional computational burden and, perhaps more importantly, their performance hinges on the performance of said generative models. Fortunately, it turns out that we can use a number of other, much simpler strategies, which we will discuss now.

@@

The most obvious and trivial mitigation strategy is to simply choose a higher decision threshold $\gamma$. This threshold determines when a proposed counterfactual should be considered as valid. Under $\gamma=0.5$, counterfactuals will end up near the decision boundary by construction. Since this is the region of maximal aleatoric uncertainty, the classifier is bound to be thrown off. By simply setting a more conservative decision threshold, we can avoid this issue to some extent. A drawback of this approach is that a classifier with high decisiveness may classify samples with high confidence even far away from the training data.

## Classifier Preserving ROAR (ClaPROAR)

Another strategy draws inspiration from ROAR @upadhyay2021robust: to preserve the classifier, we propose to simply explicitly penalize the loss it incurs when evaluated on the counterfactual $x^\prime$ at given parameter values. Recall that $\text{extcost}(\cdot)$ denotes what we had defined as the external cost in Equation \@ref(eq:collective). Then formally we let

@@

for each counterfactual $k$ where $l$ denotes the loss function used to train $M$. This approach, which we shall refer to as **ClaPROAR**, is based on the intuition that (endogenous) model shifts will be triggered by counterfactuals that increase classifier loss. It is closely linked to the idea of choosing a higher decision threshold, but likely better at avoiding the potential pitfalls associated with highly decisive classifiers. It also makes the private vs. external cost trade-off more explicit and hence manageable.

@@

Yet another strategy simply extends Wachter as follows: instead of only penalizing the distance of the individuals' counterfactual to its factual, we propose penalizing its distance to some sensible point in the target domain, for example the sample average $\bar{x}$:

@@

Once again we can putting this in the context of Equation \@ref(eq:collective): the former penalty can be thought of here as the private cost incurred by the individual, while the latter reflects the external cost incurred by other individuals. Higher choices of $\lambda_2$ relative to $\lambda_1$ will lead counterfactuals to gravitate towards the specified point $\bar{x}$ in the target domain. In the remainder of this paper, we will therefore refer to this approach as **Gravitational** generator, when we investigate its usefulness for mitigating endogenous macrodynamics^[Note that despite the naming conventions, our goal here is not to provide yet more counterfactual generators. Rather than looking at them as isolated entities, we believe and demonstrate that different approaches can be effectively combined.].

Figure \@ref(fig:mitigation) shows an illustrative example that demonstrates the differences in counterfactual outcomes when using the various mitigation strategies compared to the baseline approach, that is, Wachter with $\gamma=0.5$: choosing a higher decision threshold pushes the counterfactual a little further into the target domain; this effect is even stronger for ClaPROAR; finally, using the Gravitational generator the counterfactual ends up all the way inside the target domain in the neighbourhood of $\bar{x}$^[In order for the Gravitational generator and ClaPROAR to work as expected, one needs to ensure that counterfactual search continues, independent of the threshold probability $\gamma$.].

@@

# Generators:
generators = Dict(
    "Generic (γ=0.5)" => GenericGenerator(decision_threshold=0.5),
    "Generic (γ=0.9)" => GenericGenerator(decision_threshold=0.9),
    "Gravitational" => GravitationalGenerator(),
    "ClaPROAR" => ClaPROARGenerator()
)

@@

Our findings indicate that all three mitigation strategies are at least at par with Latent Space generators with respect to their effectiveness at mitigating domain and model shifts. Figure \@ref(fig:mitigate-results) presents a subset of the evaluation metrics for our synthetic data with overlapping classes. The top row in Figure \@ref(fig:mitigate-results) indicates that while domain shifts are of roughly the same magnitude for both Wachter and Latent Space generators, our proposed strategies effectively mitigate these shifts. ClaPROAR appears to be particularly effective, which is positively surprising, since it is designed to explicitly address model shifts, not domain shifts. As evident from the middle row in Figure \@ref(fig:mitigate-results) model shifts can also be reduced: for the deep ensemble Latent Space search yields results that are at par with the mitigation strategies, while for both the simple MLP and logistic regression our simple strategies are actually more effective. The same overall pattern can be observed for out-of-sample model performance. With respect to the other synthetic datasets, for the Moons dataset, the emerging patterns are largely the same, but the estimated model shifts are insignificant as noted earlier; the same holds for the Circles dataset, but there is no significant reduction in model performance for our neural networks; in the case of linearly separable data we find the Gravitational generator to be most effective at mitigating shifts.

@@

An interesting finding is also that the proposed strategies have a complementary effect when used in combination with Latent Space generators. In experiments we conducted on the synthetic data, the benefits of Latent Space generators were exacerbated further when using a more conservative threshold or combining it with the penalties underlying Gravitational and ClaPROAR. In Figure \@ref(fig:mitigate-latent-results) the conventional Latent Space generator with $\gamma=0.5$ serves as our baseline. Evidently, being more conservative or using one of our proposed penalties decreases the estimated domain and model shifts.

@@

Finally, Figure \@ref(fig:mitigate-real-world-results) shows the results for our real-world data. We note that for both California Housing and Credit Default data our proposed strategies do have an attenuating effect on both model shifts and performance deterioration, even though the estimated effects are somewhat less striking than for the synthetic data in Figure \@ref(fig:mitigate-results)^[Estimated domain shifts (not shown) were largely insubstantial, as in Figure \@ref(fig:real) in the previous section.]. Still, both ClaPROAR and Gravitational reduce the negative impact on out-of-sample model performance by around 25 percent from roughly 20 percentage points for the baseline approach to just 15 percentage points. For the GMSC dataset we observe no notable differences.
